Multiple visits in persons likely to have asthma

### 2018-01-12 r_scripts/analysis/asthma_care_utilization.Rmd
I have no idea how to visualize or make sense of mulitple visits.

@@

```{r setup, echo = F, warning = F, message = F}
# libraries.
library(tidyverse)

# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

>>>>

## Setup

@@

```{r setup, echo = F, warning = F, message = F}
# libraries.
library(tidyverse)
library(lme4)

# knitr options
knitr::opts_chunk$set(fig.width=8, fig.height=6, 
                      warning=FALSE, message=FALSE)
```

Reading in ZIP population-weighted smoke data and joining to ZIPCODE of persons.

```{r smoke_import_wrangle}
# read in pm2.5 data
pm_path <- "../../data/pm/2013-oregon_zip_pm25.csv"

pm <- read_csv(pm_path) %>% 
  mutate(ZIPCODE = as.character(ZIPCODE),
    geo_wt_pm = ifelse(geo_wt_pm < 0, 1, geo_wt_pm),
  # creating binary smoke indicators 
    smoke0 = ifelse(geo_smk_pm > 0, 1, 0),
    smoke5 = ifelse(geo_smk_pm > 5, 1, 0),
    smoke10 = ifelse(geo_smk_pm > 10, 1, 0),
    smoke15 = ifelse(geo_smk_pm > 15, 1, 0)) %>% 
  # create smoke wave (98%tile for 2 days) 
  group_by(ZIPCODE) %>% 
  arrange(ZIPCODE, date) %>% 
  # 98% is 15 in Oregon
  mutate(high_pm_day = ifelse(geo_wt_pm >= 15, 1, 0),
    smoke_wave = ifelse(lag(high_pm_day, oreder_by = ZIPCODE)==1 & 
                        high_pm_day==1, 1, 0)) %>% 
  rename(ZIP = ZIPCODE)
```

Reading in asthma cohort data for wildfire season and joinin to asthma cohot observations.

```{r read_data}
read_path <- paste0("./data/health/",
  "2013-oregon_asthma_fireseason_cohort.csv")

asthma_cohort <- read_csv(read_path, col_types = cols(.default = "c")) %>% 
  mutate(date = as.Date(fromdate)) %>% 
  left_join(pm, by = c("date", "ZIP"))
```

Find the person with the most observations over this timeperiod.

```{r person_most_obs}
# create indicator of number of observations
n_persons <- asthma_cohort %>% 
  group_by(personkey) %>% 
  summarise(n_obs = n())

# summary of number of persons
summary(n_persons)

# plot of distribution
ggplot(n_persons, aes(x=n_obs)) +
  geom_density()
```

92,487 unique personkey IDs, with a range of observations of 444. Most people have a reasonable number of observations over this time.

Some observations are likely multiple lines per claim. 444 visits for a single person (personkey = 10733403) over the 5 month fireseason would be highly unlikely. I want to count up the number of lines per claim per person.

```{r person_claims}
n_lines <- asthma_cohort %>% 
  group_by(personkey, clmid) %>% 
  summarise(n_lines = n())

# summary of number of lines per claim
summary(n_lines)
```

There are 265,326 unique claims. Most persons have maybe 1 to 3. One person has a unique claim with 182 (personkey = 10483893) lines associated with it. Distribution plot took too long so I didn't add it. My guess that it would show similar distribution to previous plot.

Now I want to know the number of unique visits per person.

```{r visits_per_person}
n_visits <- n_lines %>% 
  group_by(personkey) %>% 
  summarise(n_visits = n())

# summary of visits
summary(n_visits)

# density plot
ggplot(n_visits, aes(x=n_visits)) +
  geom_density()
```

Someone had 117 medical claims I'm interpreting as visits/fills over the time period (personkey = 1738926). Most people have 1 or 2 visits.

Now I'd like to know how many medical visits and pharmacy fills.

```{r claims_visits}
n_visit_types <- asthma_cohort %>% 
  # group by person and clmid
  group_by(personkey, clmid) %>% 
  # filter to first line of clmid
  arrange(personkey, clmid, line) %>% 
  # filter to only one claimline 
  filter(row_number()==1) %>% 
  # filter again to only asthma as primary or saba
  filter(visit_type != "dx_asthma_not_primary") %>% 
  # break initial grouping
  ungroup() %>% 
  # remove whitespaces from place of service
  mutate(pos_simple = stringr::str_replace_all(pos_simple, " ", "")) %>% 
  # regroup by person and service place
  group_by(personkey, pos_simple) %>% 
  # summarise
  summarise(n_type = n()) %>% 
  # spread
  spread(pos_simple, n_type) %>% 
  mutate_at(vars(Ambulance:UrgentCare), funs(ifelse(is.na(.), 0, .)))

# glipse of first couple lines
glimpse(n_visit_types)

# summary of overall
summary(n_visit_types)
```

67,738 persons. Most people have very few visits for any particular place of service (measure of severity here). I tried a couple plots, but the dataframe was too large and I don't think it will be helpful.

## Wildfire smoke

Adding in a second grouping variable, smoke > 15. Overwritting previous file.

```{r}
n_visit_types <- asthma_cohort %>% 
  # group by person and clmid
  group_by(personkey, clmid) %>% 
  # filter to first line of clmid
  arrange(personkey, clmid, line) %>% 
  # filter to only one claimline 
  filter(row_number()==1) %>% 
  # filter again to only asthma as primary or saba
  filter(visit_type != "dx_asthma_not_primary") %>% 
  # break initial grouping
  ungroup() %>% 
  # remove whitespaces from place of service
  mutate(pos_simple = stringr::str_replace_all(pos_simple, " ", "")) %>% 
  # regroup by person and service place
  group_by(personkey, pos_simple, smoke15) %>% 
  # summarise
  summarise(n_type = n()) %>% 
  # spread
  spread(pos_simple, n_type) %>% 
  mutate_at(vars(Ambulance:UrgentCare), funs(ifelse(is.na(.), 0, .)))

# checking a person that has two visits
check <- filter(n_visit_types, personkey == "10001535")
```

I think rates would be more informative than counts alone. Denominator will be number of days no smoke, and number of days smoke for each ZIP code.

I also would like to extract covariates for each person. I'll do that first.

``` {r covariates}
# extract vector of ids
ids <- unique(n_visit_types$personkey)

# extract covariates for each person
covariates <- asthma_cohort %>%
  # filter to only persons in id vector
  filter(personkey %in% ids) %>% 
  # group by personkey
  group_by(personkey) %>% 
  # sort by personkey and date
  arrange(personkey, date) %>% 
  # filter first row
  filter(row_number()==1) %>% 
  select(personkey, ZIP, gender, age) %>% 
  ungroup() %>% 
  mutate(gender = as.factor(gender),
         age = as.numeric(age))
```

```{r days_smoke}
# count up smoke days
days_smoke <- pm %>% 
  group_by(ZIP, smoke15) %>% 
  summarise(n_days=n())

# join smoke days to n_visits_types
asthma_n_visits <- n_visit_types %>% 
  # join to covariates
  left_join(covariates, by = "personkey") %>% 
  left_join(days_smoke, by = c("ZIP", "smoke15")) %>% 
  # calculate rates
  mutate_at(vars(Ambulance:UrgentCare), funs(rate10days = (./n_days)*10)) %>% 
  # filter out the NAs
  #need to return to why this might be? zipcoudes outside oregon maybe
  filter(!is.na(n_days)) %>% 
  filter(gender != "U")

summary(asthma_n_visits)

glimpse(asthma_n_visits)

```

Generalized Poisson model to see if the counts of ED visits a person might have might be greater on smoke periods compared to non-smoke periods.

```{r poisson_mod}
mod <- glmer(EmergencyRoomHospital ~ smoke15 + (1|personkey) + 
          offset(n_days), family="poisson"(link=log),
          data = asthma_n_visits)

lm_mod2 <- lmer(EmergencyRoomHospital_rate10days ~ smoke15 + (1|personkey) +
             gender + age, 
             data = asthma_n_visits)

summary(lm_mod)
summary(lm_mod2)

ggplot(asthma_n_visits, aes(x=log(EmergencyRoomHospital_rate10days*1000+0.1), 
    group = as.factor(smoke15), color=as.factor(smoke15))) +
  geom_density()
```

### Note for Sheryl: 2018-01-11

Sheryl, is this kind of what you were thinking? I have counts of numbers of times a person was seen for an asthma primary diagnosis by place of service or number of SABA fills (represented by pharmacy).

I could have expanded in different ways, but I think this would have been a huge dataframe. Also, I should note I may be undercounting because I only took the first line of unique claims.

Next step; find an indicator of smoke >10 or >15? or try smoke wave? Then compare number of visits, or rate by indicator.
